============================================================================

##Reproducible Research - Assignment 1 

============================================================================

```{r global settings, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, results="hide", warning=FALSE, message=FALSE)
```

The first step is to load and process the data for analysis.

```{r readData, echo=TRUE, results="show"}
activity <- read.csv("~/Desktop/activityunzipped/activity.csv")
```

Let's take a look at what we have loaded. 

```{r inspect, echo=TRUE, results=TRUE}
head(activity)
str(activity)
```

It appears the data variable needs to be reformatted from a factor to a date. I generally use **lubridate** to do date transformations. 

```{r loadPackage, warning=FALSE, message=FALSE}
library(lubridate)
```

```{r echo=TRUE, results="show"}
activity$date <- ymd(activity$date)
str(activity)
```

Our date is now formatted correctly. I will use the **dplyr** and **tidyr** packages to do the preprocessing. We'll use the *mutate* and *filter* functions to get a subset of the complete cases. This will be important for our aggregation work later. Finally, we will *select* only the columns date and steps for the first question.

```{r loadPackages}
library(dplyr)
library(tidyr)
```

```{r subset, echo=TRUE, results=TRUE}
actComplete <- activity %>%
  mutate(valid = ifelse(steps != "NA", 1, 0)) %>%
  filter(valid == 1) %>%
  select(steps, date)
```

Now we should be ready for the first question! 

=================================================================================

####Question 1: What is mean/median total number of steps taken per day?

=================================================================================

```{r packages}
library(dplyr)
library(tidyr)
library(reshape2)
```

First, we need to melt the dataframe into an aggregate dataset by date using the **reshape2** package.

```{r melt, echo=TRUE, results=TRUE}
dmelt <- melt(actComplete, id.vars = "date")
```

Next, we need to use **dplyr** and execute the *group_by* and *summarise* functions to get the mean/median of steps by date.

```{r aggregation, echo=TRUE, results=TRUE}
stepsDay <- dmelt %>% 
  group_by(date) %>%
  summarise(mean=mean(value), median=median(value), sum=sum(value))
```

Graph the steps per day histogram.

```{r graph1, echo=TRUE, results="show"}
hist(stepsDay$sum, main = "Histogram of Steps Per Day", xlab = "Total Steps")
rug(stepsDay$sum, col = "red")
```

Determine the mean and median total steps.

```{r sumstats, echo=TRUE, results=TRUE}
round(mean(stepsDay$sum), 2)
median(stepsDay$sum)
```

================================================================================

####Question 2: What is the average daily activity pattern?

================================================================================

This question requires a graph that plots the average steps by interval. The first step is to add back the interval variable, and filter out the missing observations.

```{r plot prep, echo=TRUE, result=TRUE}
actInterval <- activity %>%
  mutate(valid = ifelse(steps != "NA", 1, 0)) %>%
  filter(valid == 1) %>%
  select(steps, date, interval)
```

Next, we need to group the data by interval, and then average the steps by each interval.

```{r aggregation2, echo=TRUE, results=TRUE}
intervalAvg <- actInterval %>% 
  group_by(interval) %>%
  summarise(avg = mean(steps))
```

Now, we can plot the dataset above.

```{r timeseries, echo=TRUE, results=TRUE}
plot(intervalAvg$interval, intervalAvg$avg, type = "l", xlab = "Interval", ylab = "Average Steps")
```

Finally, we need to determine the maximum number of steps taken during a 5-minute interval. This is done by subsetting our dataset with a simple max command.

```{r max, echo=TRUE, results=TRUE}
intervalAvg[which(intervalAvg$avg==max(intervalAvg$avg)), ]
```

===================================================================================

####Question 3: What is the impact of imputing data for the missing observations?

===================================================================================

Determine the number of observations that have NA as its value.

```{r na, echo=TRUE, results=TRUE}
sum(is.na(activity$steps))
```

The easiest form of imputation would be to replace the NA with the mean of the interval from the other days that do not have a missing value. I will round to whole integers for uniformity of the data.

```{r impute, echo=TRUE, result=TRUE}
activity2 <- activity %>%
  mutate(step2 = ifelse(is.na(activity$step), intervalAvg$avg, activity$steps)) %>%
  mutate(impSteps = round(step2, 0)) %>%
  select(date, interval, steps, impSteps)
activity3 <- select(activity2, date, impSteps)
```

Now, we will take a look at the data with the imputation applied.

```{r result, echo=TRUE, results=TRUE}
str(activity2)
head(activity2)
tail(activity2)
```

We need prep the data for average steps by day with the new imputed dataset so I can then proceed to the graphing requirement.

```{r imputedMelt, echo=TRUE, results=TRUE}
dmelt2 <- melt(activity3, id.vars = "date")  ## create melted dataset
stepsDay2 <- dmelt2 %>%                      ## collapse by date, and summarise
  group_by(date) %>%
  summarise(mean=mean(value), median=median(value), sum=sum(value))
```

I will rerun the previous code to generate a histogram of the new imputed dataset.

```{r imputedHist, echo=TRUE, results=TRUE}
hist(stepsDay2$sum, main = "Histogram of Adjusted Steps Per Day", xlab = "Total Adjusted Steps")
rug(stepsDay2$sum, col = "blue")
```

Review the mean and median total steps of the imputed dataset to determine if there is a difference from the complete cases dataset.

```{r summarystatsimpute, echo=TRUE, results=TRUE}
round(mean(stepsDay2$sum), 2)
median(stepsDay2$sum)
```

So, it appears the missing data had very little impact when we used the mean number of steps to replace the missing values.

===========================================================================================

####Question 4: Are there differences in activity patterns between weekdays and weekends?

===========================================================================================

And ... for our final presentation, we will graph the average number of steps on weekdays and weekends. The first step is to prep the dataset by creating a factor for weekday.

```{r panelplot, echo=TRUE, results=TRUE}
activity4 <- activity2 %>%
  mutate(dayWeek = as.factor(weekdays(date))) %>%
  mutate(weekend = ifelse(dayWeek == "Saturday", 1, 
                    ifelse(dayWeek == "Sunday", 1, 0))) %>%
  select(date, interval, impSteps, dayWeek, weekend)
activity4$weekend <- as.factor(activity4$weekend)   ## convert weekend to a factor

intervalAvg2 <- activity4 %>% 
  group_by(interval, weekend) %>%
  summarise(avg = mean(impSteps))
```

Take a look at the new data to confirm that the weekend factor worked as planned.

```{r review, echo=TRUE, results=TRUE}
str(activity4)
```

The final step is craft the panel plot using **ggplot**.

```{r panelplot2, echo=TRUE, results=TRUE}
library(ggplot2)
g <- ggplot(intervalAvg2, aes(interval, avg))
g <- g + geom_line(aes(color = weekend))
g <- g + facet_grid(. ~ weekend)
g <- g + labs(x = "Interval", y = "Average Steps", title = "Weekday vs. Weekend")
g
```

It appears that the experimental subjects have a greater number of average steps on the weekends in comparison to the weekdays.